# R Fundamentals {#fundamentals}


## Package Management

Use `installed.packages() : return matrix` to list all installed package  


### Packages Listing

Below are the column names and number for filtering purpose
```{r}
dimnames( installed.packages() ) [2]
```

Perform filter based on dimension names when necessary. Set **priority = 'NA'** will filter out all base R packages  
```{r}
installed.packages( priority='NA' ) [,c(3,2)]
```

Set **priority = 'high'** will display all base R packages  
```{r}
installed.packages( priority='high' ) [,c(3,2)]
```

### Package Install / Remove

```{r eval=FALSE}
install.packages( c('ggplot', 'ggExtra'))
```

### Package Update

```{r eval=FALSE}
old.packages()           ## list all old packages, their old version and available new version
update.packages()             ## need to input y/n/c for each old package to update
update.packages( ask=FALSE )  ## automatically update all packages without asking
```


## Data Types

## Conditioinal Decision

## Looping

## Apply Family

### Overview

- Apply and its family is a useful function

### apply

### sapply

### tapply


