```{r datagen, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=2.6736, fig.height=2.5, fig.show='hold', results = 'hold') 
```

# Data Generation

Here is a review of existing methods.

## Sequential Number

### Using `:` : return vector
Produce sequantial integer with fix incremental or decremental by 1

**Incremental**
```{r}
3:6            # incremental integer
1.25:9.25      # incremental decimal
c(3:6, 4.25:8.25)  # combination of multiple sequence
```

**Decremental**
```{r}
6:3            # decremental integer
9.25: 1.25     # decremental decimal
```

### Using `seq` : return vector
Improvement from `:`, **`seq`** allows specifying **incremental steps** with `by=`.   

> `seq( from, to )`   
> `seq( from, to, by = )`   
> `seq( from, to, length.out = )     # potentially return decimal`   

**Incremental**  
```{r}
seq (3, 12)                   # default increment by=1
seq (3, 12, by = 4)           # increment of integer
seq (3.25, 12.25, by = 2.25)  # increment of decimal 
```

**Decremental** - `from` must be larger than `to`, and `by` has to be negative.  
```{r}
seq (12, 3)                    # default decrement by=-1
seq (12, 3, by = -4)           # decrement of integer
seq (12.25, 3.25, by = -2.25)  # decrement of decimal
```

**Equal Spreading** - with `length.out=`    

Equal Spreading of Integer  
```{r}
seq (10, 50, length.out = 9)   # incremental spreding of integer
seq (50, 10, length.out = 9)   # decremental spreading of integer
```

Equal Spreading of Decimal     
```{r}
seq (10.33, 50.55, length.out = 9)   # incremental spreading of decimal
seq (50.55, 10.33, length.out = 9)   # decremental spreading of decimal
```

## Random Number


### Unified Distribution

> `runif( n )                # default min=0, max=1`   
> `runif( n, min=, max= )`   

```{r}
set.seed(123)
runif(5)                 # geenrate 5 numbers within default min=0, max=1
runif(5, min=3, max=9)
```

Notice that the numbers generated are uniformly distributed.  

```{r fig.width=9, fig.height=3, fig.show='hold'}
hist(runif(300, min=3, max=9))
```

### Normal Distribution

> `rnorm( n )                # default mean=0, sd=1`   
> `rnorm( n, mean=, sd= )`   

```{r}
set.seed(123)
rnorm(5)                  # geenrate 5 numbers within default min=0, max=1
rnorm(5, mean=3, sd=1.5)
```

Notice that the numbers generated are uniformly distributed.  

```{r fig.width=9, fig.height=3, fig.show='hold'}
hist(rnorm(300, mean=3, sd=1.5))
```


### Binomial Distribution

The code below generates 'n' number of observations, each observation is the number of success for a number of trials, and there is a specific probability for success of each trial:  

The numbers generated has the below characteristic:

a. Discrete number  
b. Binomial distribution often well approximated by a **Normal distribution**, where:  
    - mean = n * prob  
    - variance = n * prob (1-prob)  
    Theoritically, when n approaches infitity, a binomial is a normal distribution  

> **`rbinorm( n, size, prob )`**   
> `rbinorm( n=, size=, prob= )`     
> $\quad$ `n = number of observations`  
> $\quad$ `size = number of trials per observations`  
> $\quad$ `prob = probability of success for each trial`  

```{r collapse=TRUE}
rbinom(100, 10, 0.4)
```

```{r}
hist(rbinom(100, 10, 0.4))
```


### Beta Distribution

**Balanced Skewness**  

```{r fig.show='hold'}
hist(rbeta(1000, 1,    1))     # no left and right skewness = uniform
hist(rbeta(1000, 10,   10))    # balanced left and right skew
hist(rbeta(1000, 100,  100))   # more balanced left and right skew
```

**Left or Right Skew**

```{r fig.show='hold'}
hist(rbeta(1000, 1000, 10))    # left skewed more
hist(rbeta(1000, 100,  1000))  # right skew more
```

### Drawing From A Bag

+ A bag has been occupied with vector `x` (produced using `:` or any other vector)  
+ `sample()` will draw from this bag  
+ Specifying `replace=T` emulate that the drwan sample will be put back to the bag for next draw    
+ R will generate error if no enough samples to draw from ( size > length(x) )  

> `sample( x=, size= )`    
> `sample( x=, size=, replace=T) # recycle number even though it is not exhausted`    

```{r results='hold'}
set.seed (123)
sample (10,   5)        # choose 5 numbers from the 'bag' containing 1:10
sample (3:10, 5)        # choose 5 numbers from the 'bag' containing 3:10

bag = runif(8, min=3, max=9)  # define the content of the bag
sample (bag, 5, replace=T)    # draw from the bag, recycling numbers
```


## Factor

**gl** generates elements based on given levels. However, it is not randomized.  

> gl ( n, k, ordered=FALSE )  
> gl ( n, k, length = n*k, ordered=FALSE )  
> $quad$ `n = number of levels`  
> $quad$ `k = number of elements to be created per level`  
> $quad$ `length = number of elements to be generated`  

```{r, collapse = TRUE, results='markup'}
gl( 3, 5, 9, c('Aaa','Bbb','Ccc'), ordered=TRUE )
str( gl(3,5, 9, c('Aaa','Bbb','Ccc')) )
```



