```{r setup_aggregation, echo=FALSE}
## global chunk option ##
knitr::opts_chunk$set(echo=TRUE, message=FALSE, fig.width=2.6736, fig.height=2.5, fig.show='hold', results='markup') 
```

# Data Summarization

## Library

|   | Package              | Function      | Purpose                                             |
|---|----------------------|---------------|-----------------------------------------------------|
| 1 | Base R               | table         | Generate frequency table                            |
|   |                      | ftable        | Generate flat table                                 |
|   |                      | prop.table    | Generate probability table out from table or ftable |
|   |                      | addmargins    | Add margin to table or ftable                       |
|   |                      | tapply        | Produce aggregation with one level of grouping      |
|   |                      | aggregate     |       |
| 2 | dplyr                | select        | select the columns             |
|   |                      | filter        | filter rows based on criteria  |
|   |                      | arrange       | sorting of rows  |
|   |                      | distinct      | unique rows only |
|   |                      | slice         | row selection    |
|   |                      | group_by      | create a tiblle table with group columns           |
|   |                      | ungroup       | remvoe group columns                               |
|   |                      | summarize     | in-group: specify the columns and functions to summarize     |
|   |                      | summarize_at  | in-group: specify the columns to be summarized               |
|   |                      | summarize_if  | in-group: conditionally choose the columns to be summarized  |
|   |                      | summarize_all | in-group: summarize all columns, be it numeric or not        |
|   |                      | top_n         | in-group: highest n rows              |
|   |                      | row_number    | in-group: automatic row numbering     |
|   |                      | n             | in-group: total number of rows        |
|   |                      | tally         | short hand for summarize(n())         |
|   |                      | count         | short hand for group_by and tally     |
| 3 | tidyr                | gather        | Gather columns into rows |
|   |                      | spread        | Reverse of gather()      |
|   |                      | seperate      | Seperate one column into multiple columns |
|   |                      | unite         | Reverse of seperate()    |

```{r,echo=FALSE}
library(dplyr)
library(tidyr)
```

## Sample Data
Sample data used simulate two categorical-alike feature, and two numeric value feature:  

- **dept** is random character between 'D1', 'D2' and 'D3'  
- **grp** is random character with randomly generated 'G1', 'G2'  
- **value1** represents numeric value, normally distributed at mean 50  
- **value2** is numeric value, normally distributed at mean 25  

```{r}
set.seed(1234)
my.df = data.frame(
  com  = paste('C',sample(1:2, 100, replace = T),sep=''),
  dept = paste('D',sample(1:3, 100, replace = T),sep=''),
  grp  = paste('G',sample(1:2, 100, replace = T),sep=''),
  team = paste('T',sample(1:2, 100, replace = T),sep=''),
  prd  = paste('P',sample(1:5, 100, replace = T),sep=''),
  value1 = rnorm(1:100, mean = 50, sd = 5),
  value2 = rnorm(1:100, mean = 20, sd = 3),
  value3 = rnorm(1:100, mean =  5, sd = 1),
  stringsAsFactors = F
)
head(my.df)
str(my.df)
```

## Frequency Table

`table` return table object (which is also array) that report frequency count base of categorical-alike data provided.  

`table` has the below data type characteristics. Note that only 2-dimensional table object is a matrix 

| Dimension | is.atomic | is.vector | is.matrix | is.array | is.table |
|:---------:|:---------:|:---------:|:---------:|:--------:|:--------:|
| 1         |     T     |     F     |     F     |     T    |     T    |
| 2         |     T     |     F     |     T     |     T    |     T    |
| 3         |     T     |     F     |     F     |     T    |     T    |
| 4         |     T     |     F     |     F     |     T    |     T    |

`ftable` is technically a **matrix** with two dimensional data (it flatten multiple dimension data). It has below data type characteristic.

| Dimension | is.atomic | is.vector | is.matrix | is.array | is.table |
|:---------:|:---------:|:---------:|:---------:|:--------:|:--------:|
| 1         |     T     |     F     |     T     |     T    |     F    |
| 2         |     T     |     F     |     T     |     T    |     F    |
| 3         |     T     |     F     |     T     |     T    |     F    |
| 4         |     T     |     F     |     T     |     T    |     F    |

### Single Dimension Data

#### Base R
Supply `table(x)` where `x` is a factor. If `x` is not factor, it will be coerce to factor  
The result is 1-D  with frequency count as number, and each category as names
```{r results='markup'}
t1 = table( my.df$com )
t1
names(t1)
is.array(t1)
```
#### dplyr

```{r results='hold'}
my.df %>% group_by(com) %>% summarize(n())
my.df %>% group_by(com) %>% tally # tally is wrapper for summarize(n())
my.df %>% count(com)  # count is a wrapper for group_by and tally
```

### Two Dimension Data

#### Base R
Supply `table(x,y)` where `x,y` are factors. If `x,y` is not factor, it will be coerce to factor  
Frequency table for two dimensional variables is a  **2-D matrix/array**
```{r results='hold'}
t2 = table( my.df$com, my.df$dept )
t2
is.matrix(t2)
dim(t2)
colnames(t2)
rownames(t2)
```

#### dplyr with tidyr
Supply `dplyr::group_by` with two dimensional variables. Note that `count()` is a short form for `group_by()` and `summarize()`. However, it is **not possibel to rename column n** using `count()`
```{r results='markup'}
my.df %>% group_by(com,dept) %>% tally
my.df %>% count(com,dept) # count is wrapper for group_by and tally
my.df %>% count(com,dept) %>% spread(dept,n) # spread com into column
```

### Three Dimension Data
#### Base R
3-D frequency table is a 3-D array.
For3-D table, 2-D table is presented for each 3rd dimension data
```{r results='markup'}
t3 = table( my.df$com, my.df$dept, my.df$grp )  #last dimension is grp
dim(t3)
t3
rownames(t3)
colnames(t3)
```

Alternatively, use `ftable()` to flatten 3-D table above. **Last dimension** is always presented at column
```{r results='markup'}
f3 = ftable( t3 )
f3
str(f3)
```
#### dplyr
The result of dplyr on multi-level of grouping is similar to flatten table using `ftable()`
```{r results='markup'}
my.df %>% group_by(com, dept, grp ) %>% tally
my.df %>% group_by(com, dept, grp ) %>% tally %>% spread(grp,n)
```

### Four Dimension Data
When table contain three or more dimension, use `ftable (flat table)` to put multi dimension table into one flat output. **Last dimension** is always presented at column
```{r results='markup', class.source='bg-success', class.output='bg-info'}
t4 = table( my.df$com, my.df$dept, my.df$grp, my.df$team )  # last dimension is team
f4 = ftable( t4 )
f4
```
#### dplyr
The result of dplyr on multi-level of grouping is similar to flatten table using ftable()
```{r results='markup'}
my.df %>% group_by(com, dept, grp, team ) %>% summarize(freq=n())
```
### Making Table Proportion

`prop.table` **converts** `table` or `ftable` object into proportion. It can calculate table-wise, column-wise or row-wise proportion.  

> prop.table (x,margin=NULL)  
> $\quad$ x = table object  
> $\quad$ margin = NULL: proportion table-wise, 1-row_wise, 2-column_wise  

#### Proportion Table on 'table' object  

```{r results='markup'}
prop.table( t1 )
prop.table( t2 )
prop.table( t2, margin=1 )
prop.table( t2, margin=2 )
```

#### Proportion Table on 'ftable' object  

```{r results='markup'}
prop.table( f3 )
prop.table( f4 )

prop.table( f3, margin=1 )
prop.table( f4, margin=1 )

prop.table( f3, margin=2 )
prop.table( f4, margin=2 )
```

### Adding Margin Info To Table

> addmargins (x, margin=NULL)  
> $\quad$ x = table or ftable object  
> $\quad$ margin = NULL: row and column-sum, 1-col_sum, 2-row_sum  

#### Margin Info on 'table' object
```{r results='markup'}
addmargins( t2)
addmargins( t2,margin=1 )
addmargins( t2,margin=2 )
addmargins( t3 )
addmargins( t4 )
```

#### Margin Info on 'ftable' object
```{r results='markup'}
addmargins( f3 )
addmargins( f4 )
```

### Proportion Table with Margin

First to obtain the proportion table, then only add the margin.  

```{r results='markup'}
addmargins( prop.table( t2 ))    # add both column and row margin
addmargins( prop.table( t2 ), 1) # add column margin
addmargins( prop.table( t2 ), 2) # add row margin
addmargins( prop.table( t3 ))
addmargins( prop.table( t4 ))
```

```{r results='markup'}
addmargins( prop.table( f3 ))
addmargins( prop.table( f4 ))
```

## Data Aggregation

This chapter explore multiple methods to group data and computes value within groups:  

1. tapply
2. aggregate (10x slower than apply)
3. dplyr and tidyr, a much intuitive tools

### Single Dimension

#### Base R

##### tapply

> **`tapply ( X, INDEX, FUN, na.rm = FALSE )`**  
> $\quad$ ` X = value vector`  
> $\quad$ ` INDEX = vector of groups data (can be factor, number or string)`  
> $\quad$ ` FUN = function to apply on X, according to group specified in INDEX`  
> $\quad$ ` na.rm = ignore <NA> values`  

- `X` and `INDEX` must have the same length  
- `tapply` will divide the value vector (X) into groups (base on INDEX), and perform computation (FUN) on each group  
- If there are **`<NA>`** in the vector X, some FUN may fail to return value, such as mean, sum.  So it is essential to specify **`na.rm = TRUE`** in these cases  
- Group identifier will be used as column name (accessible through **`names()`**)  

```{r results='hold'}
t1 = tapply(my.df$value1, my.df$com, mean)
t1
names( t1 )
```

**FUN That Returns Non-Vector -  output is list**

```{r results='markup'}
t2 = tapply(my.df$value1, my.df$com, FUN=summary)
t2
names( t2 )
```

#### dplyr

```{r results='markup'}
my.df %>% group_by(com) %>% summarize(v1_avg=mean(value1))
```
### Two Dimension

#### Base R
##### tapply
Wrap multiple categorical alike variables in a list for multi level grouping
```{r results='hold'}
t3 = tapply(my.df$value1, list(my.df$com, my.df$dept), mean)
t3
rownames(t3)
colnames(t3)
dim(t3)
is.matrix(t3)
```
##### dplyr
```{r results='hold'}
my.df %>% group_by(com,dept) %>% summarize(v1_avg=mean(value1))
my.df %>% group_by(com,dept) %>% summarize(v1_avg=mean(value1)) %>% spread(dept,v1_avg)
```

### Three Dimension

#### Base R
##### tapply
```{r results='hold'}
t4 = tapply(my.df$value1, list(my.df$com, my.df$dept, my.df$grp), mean)
t4
rownames(t4)
colnames(t4)
dim(t4)
is.matrix(t4)
```
#### dplyr
```{r results='hold'}
my.df %>% group_by(com,dept,grp) %>% summarize(v1_avg=mean(value1))
my.df %>% group_by(com,dept,grp) %>% summarize(v1_avg=mean(value1)) %>% spread(dept,v1_avg)
```



### aggretate - base R

**Aggregate** is a very useful base R function and provides quick way to group data and values:  

- Input in **list/data.frame**, computes and **output new data.frame**.  
- It groups categorical variable(s) and compute value variable(s) based on function FUN.  
- FUN can be **min, max, mean, sd, sum or length (frequency count)**.  
- **ONLY ONE function** is supported, and it applies to all value variables !!!

#### Basic Syntax (formula method) - data source is data.frame
The **formula method** use **'data'** parameter and therefore apply for single data source only. The objective is simplicy and without flexibility to customize column names

> **`aggregate (data = df, formula, FUN = function)  `**  
> $\quad$ Formula in the form: **value~categorical**   

#### One value variable ~ one categorical variable
```{r}
aggregate (data = my.df, value1 ~ grp, FUN = length)
```

#### One value variable ~ multiple categorical variables
```{r}
aggregate (data = my.df, value1 ~ grp + dept, FUN = length)
```

#### Multiple value variables ~ one categorical variable, use cbind()
```{r}
aggregate (data = my.df, cbind(value1,value2) ~ grp, FUN = length)
```

#### multiple value variables ~ multiple categorical variable
```{r}
aggregate (data = my.df, cbind(value1,value2) ~ grp + dept, FUN = length)
```

#### ALL value variables ~ multiple categorical variable, use dot notation
Change from FUN=length to sum results in **error** because sum() cannot be applied to non-numerical variable 'team'
```{r}
aggregate (data = my.df, . ~ grp + dept, FUN = length)
```

#### Advance Syntax (by method) - data source is either list or data.frame

The **advantage** of '**by method'** are:

- Can use list/data.frame subset method to choose column to display, hence flexible  
- Can customize output column names (list subset method only)  
- Flexibility to use multiple data sources, hence **'data' is not used** and has no effect if specified  

> Using **list subseting:** column name is not preserved, hence must specify meaningful column names. If not supplied, generic names and undesirable column names derived from data value will be used as column name  
_**aggregate (x = list(...value_variables...), by = list(...categorical_variables...), FUN = function)**_


```{r}
aggregate (x =  list( v1_mean = my.df$value1, my.df$value2 ), 
           by = list( my.df$grp, DEPT = my.df$dept), FUN=mean)
```

> Using **data.frame subseting:** column names are preserved and no option to change. Notice attempt below to change the column name does not succeed  
_***aggregate( x = df[,c(...)], by = df[,c(...)]), FUN = function) ***_  
_***aggregate( x = df[, p:q], by = df[,s:t]), FUN = function) ***_
```{r}
aggregate(x=my.df[, c(v1_mean='value1', 'value2')], by=my.df[,c(GRP='grp', 'dept')], FUN=mean)
# aggregate(x = my.df[, 4:5], by = my.df[, 1:2], FUN = mean) # produce similar result as above
```

### dplyr

#### Grouping
- Grouping returns all rows, with **internal grouping structure**  
- Display of tibble rows **truncates automatically** if too many rows. Use `print(tbl_df(my.tbl), n=40)` to display more rows (quite troublesome, but look at all the other benefits)  
- The internal grouping structure can be used for **subsequent summarization**  
- To get the row numbers for each group, add additional column with **mutate** of **row_number()**  
```{r}
d.df = my.df %>%
  group_by(com,dept, grp) %>%
  arrange(com,dept,grp) %>%
  mutate(idx = row_number())

print(tbl_df(d.df),n=18)
```
Verify group levels in `attr(*, "vars")`
```{r}
attr(d.df, 'vars')
```

#### Frequency Counting

**`count()`** groups and provide additional column **`n`**, which counts the number of rows within the group 
```{r, result='hold'}
my.df %>% count(com, dept, prd)
```

**`count()` is shorthand for group_by and summarize**
```{r, result='hold'}
my.df %>% group_by(com, dept, prd) %>% summarize(count = n())
```

**`summarize()` reduce grouping level by one**
```{r, results='hold'}
my.df %>% group_by(com, dept, grp)%>% attr('vars')
my.df %>% group_by(com, dept, grp) %>% summarize(n()) %>% attr('vars')
```

**BUT** `count()` reset grouping entirely (no grouping) !!!
```{r, result='hold'}
my.df %>% count(com, dept, grp) %>% attr('vars')
```

#### Summarization

##### summarize: Apply Multiple Functions to Multiple Columns
Different functions can be applied to different variables, each return a new column
```{r}
my.df %>%
  group_by(com, dept, prd) %>%
  summarise( avg_value1 = mean(value1),
             min_value2 = min(value2),
             max_value3 = max(value3),
             n.team = n_distinct(team),
             count = n())
```

##### summarize_at: Apply Function(s) To Multiple Columns

> **`summarize_at ( vars(...), funs(...) ), ...`**   

> $\quad$ ` Use vars() to specify multiple variables: vars(7:9), vars(7,8,9), vars('col1','col3','col5'), vars(col1:col3)`  
> $\quad$ ` Use funs() to specify multiple functions: funs(func1,func2,func3)`  
> $\quad$ ` ... : additional parameters to be supplied to funs`  

Pick variables by their **names**
```{r}
my.df %>%
  group_by(com, dept, prd) %>%
  summarise_at(vars('value2', 'value3'), mean, na.rm=T)
```
Pick variables by their **index** number
```{r}
my.df %>%
  group_by(com, dept, prd) %>%
  summarise_at(vars(7,8), mean)
```
PIck **range** of variables by **column name range**
```{r results='hold'}
my.df %>% group_by(com, dept, prd) %>% 
  summarise_at(vars(value2:value3),mean) 
```
Pick **range** of variables by **column index range**
```{r results='hold'}
my.df %>% group_by(com, dept, prd) %>% 
  summarise_at(vars(7:8),mean) 
```
**Multiple Functions to Multiple Columns**
```{r results='hold'}
my.df %>% group_by(com, dept, prd) %>% 
  summarise_at(vars('value1','value2'),funs(mean,sd), na.rm=T)
```

##### summarize_all: Apply Function(s) to ALL Columns
This apply function(s) to all non-grouped variables  

> **`summarize_all (  funs(...) ), ...`**  
> $\quad$ ` Use funs() to specify multiple functions: funs(func1,func2,func3)`  
> $\quad$ ` ... : additional parameters to be supplied to funs`  

Notice that each column whin the group will be summarized with one more functions. This includes columns which are NOT numeric, resulting warnings and NA.  

**Single Function to All Variables**  
If only one function, `funs()` is not required.  
Notice below non-grouped variable **prd** has been 'mean' but produced NA since it is non-numeric
```{r,results='hold' }
my.df %>%
  group_by(com, dept, grp, team) %>%
  summarise_all(mean, na.rm=T)
```

**Multiple Functions to All Variables**  
Use `funs()` to supply multiple functions
```{r,results='hold' }
my.df %>%
  group_by(com, dept, grp,team, prd) %>%
  summarise_all(funs(mean,sd), na.rm=T)
```

##### summarize_if: Apply Function(s) to All Matching Columns
**Single Function To All Matching Columns**
```{r,results='hold' }
my.df %>%
  group_by(com, dept) %>%
  summarise_if(is.numeric, mean, na.rm=T)
```

**Multiple Functions To All Matching Columns**
Any of the above summarize can combine with  **`funs()`** to specify multiple functions.  
```{r,results='hold' }
my.df %>%
  group_by(com, dept, prd) %>%
  summarise_if(is.numeric, funs(mean,sd))
```

##### Summarize Reduce The Grouping Level by One
- Each call to `group_by()` will reset group level
- Each call to `summarize()` will reduce group level by one
```{r,results='hold' }
my.df %>% group_by(com,dept,grp) %>% attr('vars') 
my.df %>% group_by(com,dept,grp) %>% summarize(n()) %>% attr('vars') 
my.df %>% group_by(com,dept,grp) %>% summarize(n()) %>% summarize(n()) %>% attr('vars') 
```
#### In-Group Row Numbering
- Summarize reduce group level by one, hence n() will return the rows within the new group.  
- Example below group with three levels (com, dept,prd), however, after summarize, n() returns the number of rows within (com,dept) group.  
- **`row_number()`** dynamically assign **sequential row number within the group**
```{r, result='hold'}
my.df %>%
  group_by(com,dept) %>%
  summarize(avg.value1=mean(value1)) %>%
  mutate (rnum = row_number())
```

Alternatively, **mutate with `1:n()`** will return the same result as above. `n()` is the total number of rows within group, `1:n()` will return a sequential integer number
```{r, result='hold'}
my.df %>%
  group_by(com,dept) %>%
  summarize(avg.value1=mean(value1)) %>%
  mutate (rnum = 1:n())
```

#### In-Group Row Selection

##### Top-N Selection
Example below select highest occurance of  **prd** within the group of (com,dept).  It is based on the principe that count reduce group levels by one, then selection is based on rows within the new group  

> **`top_n (n,wt)`**  
> $\quad$ ` n : number of rows to return`  
> $\quad$ ` wt: column to use, default order using last column`  

```{r}
my.df %>% group_by(com,dept,prd) %>% summarize(t1=sum(value1), t2=sum(value2))
my.df %>% group_by(com,dept,prd) %>% summarize(t1=sum(value1), t2=sum(value2)) %>% top_n(n=1, wt=t2)
```
## R with SQL Emulation

Running SQL statement on existing data.frame are useful to derive summarization and aggregation for someone who are familiar with SQL.  

### Library

**sqldf** library is required.  It has dependency on gsubfn, proto and RSQLite packages.  

```{r}
library(sqldf)
```

### Run The Code

R data.frame variable is specified in **'FROM'** clause.  
Note that `.` is a SQL operator, any variable with '.' must be contained within **single quote**.  

```{r}
sqldf("
  SELECT com, dept, count(*) AS qty, AVG(value1) AS v1_mean, SUM(value2) AS v2_sum
  FROM 'my.df'  
  GROUP BY com, dept
")
```
