# Data Preprocessing

## Artificial Grouping

Artificial group can be created based on existing numeric data.  Such as age_group based on age.  

### Grouping with Numeric Breakpoints

Simulate data with x,y,z variables. p simulates priority scoring.  

```{r}
x = rnorm(10, mean = 10)
y = rnorm(10, mean = 10)
p = x * y
```

Articial groups is created first by identifying the **number of groups**, generate the break points vector, then **cut** the data base on **break points** and return **factor** as output.  

Automatically calculate breakpoints by distributing numbers into the min-max range, in low to high order:  

```{r, results='markup'}
num_groups = 4
breakPoints = seq(min(p), max(p), length.out = num_groups + 1)
breakPoints
```

> `cut ( x, breaks, right = TRUE, include.lowest = FALSE)`  
> $\quad$ `x: numeric vector to be cutted`   
> $\quad$ `breaks: numeric vector ranging from low to high (in order)`  
> $\quad$ `include.lowest: FALSE - ommit element matching lowest number in breaks`  
> $\quad$ `right: TRUE - close end on right;open end on left`  

The result from **cut** is factor based on order from breakPoints. Therefore, once convert into numeric, the group number is in order of low to high accoriding to breakPoints.  

Verify that group (g) has been assigned for each priority (p).    

```{r}
g = as.numeric( cut( p, breakPoints, include.lowest=TRUE))
data.frame(p,g)
```

### Grouping based on Custom Criteria

???

## Handling Missing Data

### Detecting Missing Data

### Removing Missing Data


## Scaling Data

### Sample Data

Scaling section will use sample data generated as below:  
```{r}
set.seed(1234)
my.df = data.frame(
  id = paste('ID_', 1:5, sep = ''),
  value1 = sample(50:100, 5),
  value2 = sample(10:50, 5),
  stringsAsFactors = F
)
my.df
```

### Z-Score Scaling

`scale` apply transformation **column-wise**, for  columns within matrix or dataframe  
`scale` return a **matrix**  

>`scale (x, center=T, scale=T)`  # default S-Score transformation  
>$\quad$ `center = T (default) means value minus with mean`  
>$\quad$ `scale  = T (default) means value divide by sd`  
>$\quad$ $\quad$ `output scaled:center --> mean`  
>$\quad$ $\quad$ `output scaled:scale  --> sd`  

```{r results='markdown'}
scale( my.df[,2:3] )
scale( my.df[,2:3], scale=F )
```

### Min Max Scaling

Min-Max scaling will transform all numbers between 0 and 1.  
Easiest way to perform this transformation is to write a function then `apply`.  

```{r}
min_max = function(x){(x-min(x))/(max(x)-min(x))}
apply( my.df[,2:3], MARGIN = 2, FUN = min_max )  ## 2 means column-wise
```


