# Data Preprocessing


## Handling Missing Data

### Detecting Missing Data

### Removing Missing Data


## Scaling Data

### Sample Data

Scaling section will use sample data generated as below:  
```{r}
set.seed(1234)
my.df = data.frame(
  id = paste('ID_', 1:5, sep = ''),
  value1 = sample(50:100, 5),
  value2 = sample(10:50, 5),
  stringsAsFactors = F
)
my.df
```

### Z-Score Scaling

`scale` apply transformation **column-wise**, for  columns within matrix or dataframe  
`scale` return a **matrix**  

>`scale (x, center=T, scale=T)`  # default S-Score transformation  
>$\quad$ `center = T (default) means value minus with mean`  
>$\quad$ `scale  = T (default) means value divide by sd`  
>$\quad$ $\quad$ `output scaled:center --> mean`  
>$\quad$ $\quad$ `output scaled:scale  --> sd`  

```{r results='markdown'}
scale( my.df[,2:3] )
scale( my.df[,2:3], scale=F )
```

### Min Max Scaling

Min-Max scaling will transform all numbers between 0 and 1.  
Easiest way to perform this transformation is to write a function then `apply`.  

```{r}
min_max = function(x){(x-min(x))/(max(x)-min(x))}
apply( my.df[,2:3], MARGIN = 2, FUN = min_max )  ## 2 means column-wise
```


