# Find, Order and Filter Data

## Sample Data
Sample data used simulate two categorical-alike feature, and two numeric value feature:  

- **dept** is random character between 'D1', 'D2' and 'D3'  
- **grp** is random character with randomly generated 'G1', 'G2'  
- **value1** represents numeric value, normally distributed at mean 50  
- **value2** is numeric value, normally distributed at mean 25  

```{r}
set.seed(1234)
my.df = data.frame(
  dept = sample(1:3, 10, replace = T),
  value1 = rnorm(1:10, mean = 50, sd = 5),
  stringsAsFactors = F
)

my.vector = sample(1:100,18, rep=T)

head(my.df)
head(my.vector)
```

## Finding Data

Find the **first match row number(s)** of specific element(s).  

> `match ( x, y )`  
> $\quad$ `x = vector of criteria`  
> $\quad$ `y = vector of elements to find from`   

```{r}
match(2, my.df$dept)
match(c(1,2), my.df$dept)
```

## Ordering Data

The key **idea** of ordering data is to produce an 'order' vector representing the **position** of the element.  Then apply the order list to vector/data.frame to produce sorted result.  

**order** make ordering based on a **numeric vector**. The order result can be applied to vector or data.frame.  

> **`order`** `( x,             x must be numeric number`  
> $\quad$ `decreasing = FALSE,   ascending or descending`  
> $\quad$ `na.last = TRUE)       if TRUE, NA value are put last`  

### Ordering Vector

Create the order.

```{r}
the.order = order (my.vector, decreasing=T) 
the.order
```

Apply the order on variable.  

```{r}
the.order = order (my.vector, decreasing=T) 
the.order
my.vector[ the.order ]
```

### Ordering Data Frame

**One Level Ordering**  
```{r}
the.order = order(my.df$dept, decreasing=TRUE)
my.df[the.order,]
```

**Multi Level Ordering**  

Specify more than one column for multi-level ordering.However, all levels follow the same decreasing (ascending/desceding).  

```{r results='hold'}
attach(my.df)
the.order = order(dept, value1, decreasing=TRUE)
my.df[the.order,]
detach(my.df)
```


## Filtering Data

There are two methods of filtering data, using logical vector or subset() function.  

With both methods, **row.names are retained** in the output vector/dataframe.  

- **subset()** is a general function that can be used to filter vector, matrix and data.frame  
- logical vector method - derive a conditional criteria that produce a logical vector, then apply to element selection
- `which()` takes logical vector and return actual indices of TRUE elements. The output from which can be use for subsetting  

### Subseting Vector

All methods show in below section has similar results.  

#### Using subset()

```{r}
my.vector = 1:100
subset( my.vector, my.vector>10 & my.vector<20)
```

#### Using logical vector

Essentially, the below has similar results as using subset()..  

```{r}
my.vector = 1:100
lvector = my.vector>10 & my.vector<20
my.vector[ lvector ]
```

#### Using row numbers vector

**`which`** returns the **true indices of a logical object**.  

```{r}
my.vector = 1:100
which (my.vector>10 & my.vector<20)
my.vector[which (my.vector>10 & my.vector>10& my.vector<20)]
```

**'which'** has a useful form to return row number with max or min value  

```{r}
my.vector = 1:100
which.min (my.vector)
my.vector[which.min(my.vector)]

which.max (my.vector)
my.vector[which.max(my.vector)]
```


### Subseting Data Frame

All methods show in below section has similar results.  

#### Using subset()

```{r}
my.df = data.frame( 
  x = 1:100,
  y = 300:201
)
subset( my.df, x<10 & y<300 )
```

#### Using logical vector

```{r}
lvector = my.df$x<10 & my.df$y<300
my.df [ lvector, ]
```

#### Using row numbers vector

**`which`** returns the **true indices of a logical object**.  

```{r}
z = which (my.df$x<10 & my.df$y<300)
my.df [ z, ]
```
