```{r ordering, echo=FALSE}
## global chunk option ##
knitr::opts_chunk$set(echo=TRUE, message=FALSE, warning=FALSE, fig.width=2.6736, fig.height=2.5, fig.show='hold', results = 'hold') 
```

# Find, Order and Filter Data

## Library

```{r}
library(dplyr)
```

Functions and libraries used in this chapter are as below:  

|   | Package              | Function      | Purpose                                                                                                 |
|---|----------------------|---------------|---------------------------------------------------------------------------------------------------------|
| 1 | Base R               | match         | return the position(s) of first match                                                                   |
|   |                      | order         | return the positions for each vector element with order                                                 |
|   |                      | which         | return the positions of value TRUE of a logical vector                                                  |
|   |                      | subset        | return a new data.frame that match the criterias                                                        |
| 2 | dplyr                | filter        | return a new data.frame that match the criterias                                                        |
|   |                      | arrange       | return a new data.frame that match the criterias                                                        |

## Sample Data

```{r echo=FALSE}
set.seed(1234)
my.df = data.frame(
  dept = paste('D',sample(1:3, 10, replace = T),sep=''),
  value1 = rnorm(1:10, mean = 50, sd = 5),
  stringsAsFactors = F )

my.vector = sample(1:100,17, rep=T)
```

### Sample Data Frame

```{r  collapse=T}
str( my.df )

head( my.df )
```


### Sample Vector

```{r results='markup', collapse=T}
str( vector )

head( my.vector )
```

## Finding Data in Vector

Find the **first match position number(s)** of specific element(s).  

> `match ( x, y )`  
> $\quad$ `x = vector of criteria`  
> $\quad$ `y = vector of elements to look in`   

Let's look at the **sample vector**.  

```{r, collapse=TRUE}
my.df$dept
```

Find the position of **one criteria**.  

```{r, collapse=TRUE}
match( 'D2', my.df$dept )
```

Find the positions of **multiple criterias**.  

```{r, collapse=TRUE}
match( c('D1','D2','D3'), my.df$dept )
```


## Ordering Data

The key **idea** of ordering data is to produce an 'order' vector representing the **position** of the elements.  Then apply the order list to vector/data.frame to produce sorted result.  

**order** make ordering based on a **numeric vector**. The order result can be applied to vector or data.frame.  

> **`order`** `( x,             x must be numeric number`  
> $\quad$ `decreasing = FALSE,   ascending or descending`  
> $\quad$ `na.last = TRUE)       if TRUE, NA value are put last`  

### Ordering Vector

Let's look at the **sample vector**.  

```{r, collapse=TRUE}
my.vector
```

**Create the order**.  

```{r, collapse=TRUE}
the.order = order (my.vector, decreasing=T) 
the.order
```

**Apply the order** on vector.  

```{r, collapse=TRUE}
my.vector[ the.order ]
```

### Ordering Data Frame

We want to order a data.frame based on the values in one or more columns.  
Here is the data.frame example.  

```{r, collapse=TRUE}
head(my.df)
```

#### One Level Ordering

Order the data.frame based on one column - dept.  

```{r}
the.order = order(my.df$dept, decreasing=TRUE)
head( my.df[the.order,] )
```

#### Multi Levels Ordering

Order the data.frame based on two columns. **However, all columns follow the same decreasing** (ascending/desceding).  

```{r results='hold'}
attach(my.df)
the.order = order(dept, value1, decreasing=TRUE)
head( my.df[the.order,] )
detach(my.df)
```

#### dplyr::arrange

- **dplyr::arrange** offers the best flexibility and simplicity  
- It overcomes base **order()** with ability to order multiple columns with different ascending/descending mode  
- It assume all ascending ordering unless **desc** is specified  

```{r}
arrange(my.df, dept, desc(value1))
# my.df %>% arrnage(dept, desc(value1))
```

## Filtering Data

There are multiple methods of filtering data:  

- Using **logical vector**  
- Using **`subset()`** function  
- Using row numbers: **`which()`**  
- Using **`dplyr::filter()`** function  

With these methods, **row.names are retained** in the output vector/dataframe.  

- **subset()** is a general function that can be used to filter vector, matrix and data.frame  
- logical vector method - derive a conditional criteria that produce a logical vector, then apply to element selection
- `which()` takes logical vector and return actual indices of TRUE elements. The output from which can be use for subsetting  

### Subseting Vector

All methods show in below section has similar results.  

#### Using subset()

```{r}
my.vector = 1:100
subset( my.vector, my.vector>10 & my.vector<20)
```

#### Using Logical Vector

**First, create a logical vector**.  

```{r, collapse=T}
my.vector = 1:100
lvector = my.vector>10 & my.vector<20
head( lvector, 24 )
```

**Then apply the logical vector with selection**.  

```{r, collapse=T}
my.vector[ lvector ]
```

#### Using Row Numbers

**`which`** returns the **indices of TRUE value, of a logical vector**.  

**Create a vector containing row numbers for all TRUE values**.  

```{r, collapse = TRUE}
my.vector = 1:100
w = which (my.vector>10 & my.vector<20)
w
```

**Then apply the row numbers with selection**.  

```{r, collapse=TRUE}
my.vector[ w ]
```

**'which'** has a useful form to return position of **max or min** value.  

```{r, collapse=TRUE}
my.vector = 1:100
which.min ( my.vector )

my.vector [ which.min(my.vector) ]
```

```{r, collapse=TRUE}
which.max ( my.vector )

my.vector [ which.max(my.vector) ]
```

### Subseting Data Frame

Subsetting data.frame is generally similar to subsetting vector, except that it uses **[rows,cols] selectors**.  

#### Using subset()

```{r}
my.df = data.frame( 
  x = 1:100,
  y = 300:201
)
head( my.df ) 

subset( my.df, x<10 & y<297 )
```

#### Using Logical Vector

```{r}
lvector = my.df$x<10 & my.df$y<297
my.df [ lvector, ]
```

#### Using Row Numbers

**`which`** returns the **positions of value TRUE of a logical vector**.  

```{r}
z = which (my.df$x<10 & my.df$y<297)
my.df [ z, ]
```


#### Using dplyr::filter()

- Logical comparison operator '&, |' can be used, just like **subset()**  
- Multiple filter criterias seperate with ',' are considered as '&' condition  

```{r, collapse=TRUE}
filter(my.df, x<10, y<297)

filter(my.df, x<10 & y<297)
```

**Elegantly**, dplyr powerful **%>%** comes into play:  

```{r}
my.df %>% filter( x<10, y<297)
```
