```{r init}
## global chunk option ##
knitr::opts_chunk$set(echo=TRUE, message=FALSE, fig.width=2.6736, fig.height=2.5, fig.show='hold') 
```

# Graphic Visualization

This chapter compares various method to plotting using base-R and ggplot.

## Library used

- Loading necessary library as below:
- Base R library already included functions: ** hist, plot, barplot, boxplot**

```{r}
library(ggplot2)    ## ggplot, qplot
```


## Sample Data
This chapter uses the sample data generate with below code. The idea is to simulate two categorical-alike feature, and two numeric value feature:  

- **dept** is random character between 'D1', 'D2', 'D3', 'D4' and 'D5'  
- **grp** is random character with randomly generated 'G1', 'G2'  
- **value1** represents numeric value, normally distributed at mean 50  
- **value2** is numeric value, normally distributed at mean 25  

```{r}
set.seed(1234)
my.df = data.frame(
  dept = paste('D',sample(1:5, 100, replace = T),sep=''),
  grp  = paste('G',sample(1:2, 100, replace = T),sep=''),
  value1 = rnorm(1:100, mean = 50, sd = 5),
  value2 = rnorm(1:100, mean = 20, sd = 3),
  stringsAsFactors = F
)
head(my.df)
```

## Histogram

### Single Dimension Data
- Require **x** as numerical data
- In **hist**, **binwidth** setting is not available, only breaks (number of bins) can be specified
- Default hist/ggplot/qplot number of bins is **30**
- In **qplot**, single x numerical variable **default to histogram**
- You can't specify both **bins/breaks** and **bindwidth** together, as it implies each other

```{r, fig.width=9}
par(mfrow=c(1,2))
hist  (my.df$value1) # default breaks = 30
hist  (my.df$value1, breaks=3)
```

```{r}
qplot (data = my.df, x=value1)
qplot (data = my.df, x=value1, geom='histogram')
qplot (data = my.df, x=value1, bins=15)
ggplot(data = my.df, aes(x=value1)) + geom_histogram() # default bins = 30
ggplot(data = my.df, aes(x=value1)) + geom_histogram(bins = 15)
ggplot(data = my.df, aes(x=value1)) + geom_histogram(binwidth = 10)
```


### Two Dimension Data
- x = numerical data  
- fill = categorica-alike data
```{r, fig.width=4}
qplot (data = my.df, x=value1,  fill=grp, geom='histogram')
ggplot(data = my.df, aes(x=value1, fill=grp)) + geom_histogram()
```


## Scatter Plot
### Two Dimension Data
- Use scatter plot to represent **correlation** between two numeric variables
- x = number, y = number  
- qplot default to **geom_point** when two numerical value is supplied for x and y

```{r, fig.width=8}
plot  (my.df$value1, my.df$value2)
```

```{r}
qplot (data = my.df, x = value1, y = value2)
qplot (data = my.df, x = value1, y = value2, geom='point')
ggplot(data = my.df, aes(x=value1, y=value2)) + geom_point()
```


### Two + One Dimension Data
- Base-R does not support extra dimension visualization
- In **qplot/ggplot**, the third dimension of data can be represented by assigning **color** parameter to the third variable
- Note that **fill has not effect** on scatter plot. **fill** should only be used for bar like chart eg. **geom_hist** or **gem_bar**

```{r, fig.width=8}
plot  (my.df$value1, my.df$value2)
```

```{r}
qplot (data = my.df, x = value1, y = value2, color = grp, geom='point')
ggplot(data = my.df, aes(x=value1, y=value2, color = grp)) + geom_point()
ggplot(data = my.df, aes(x=value1, y=value2, fill = grp)) + geom_point()
```


## Bar Chart

### Single Dimension Data
- Use bar to repfresent **frequency** chart
- **plot** requre a factor to plot frequency chart
- **barplot** require conversion of vector into **table** for plotting
- **qplot** default to **geom_bar** when **single categorical-alike** feature is used

```{r,fig.width=8}
par(mfrow=c(1,2))
plot(as.factor(my.df$dept))
barplot(table(my.df$dept))
```

```{r}
qplot (data = my.df, x=dept)
qplot (data = my.df, x=dept, geom='bar')
ggplot(data = my.df, aes(x=dept)) + geom_bar()
```

### Two + One Dimension Data
- Use **fill**  to introduce extra variable visualizion in filling the bar
- Use **color** to have the extra variable represented with **border** color
```{r}
qplot (data = my.df, dept, fill = grp)
qplot (data = my.df, x = dept, fill = grp, geom='bar')
ggplot(data = my.df, aes(x = dept, fill = grp)) + geom_bar()
ggplot(data = my.df, aes(x = dept, color= grp)) + geom_bar()
```

### Reordering

```{r,fig.width=8}

```

```{r}
qplot (data = my.df, x=dept)
qplot (data = my.df, x=dept, geom='bar')
ggplot(data = my.df, aes(x=dept)) + geom_bar()
```


### Positioning
- **qplot does not** support positioning
- For **ggplot/qplot**, default position is **stack**
- position = 'dodge' similar to **position = position_dodge()**, however the later is more flexible with ability to adjust overlapping level between sub-bar (default is 0.9)

```{r}
g = ggplot(data = my.df, aes(x=dept, fill=grp)) 
g + geom_bar(position='stack') # default position
g + geom_bar(position='dodge')
g + geom_bar(position=position_dodge()) # default 0.9
g + geom_bar(position=position_dodge(0.5))
g + geom_bar(position=position_dodge(1.0))
```

### In-Bar Text Labeling


## Box Plot

### One Dimension Data

- In boxplot(), only single variable need to be supplied
- In **ggplot/qplot**, variable x and y is required. Variable y is the actual value, variable x is the group variable. Case of one dimension, use x='' when no grouping is desired

```{r}
boxplot(my.df$value1)
qplot  (data = my.df, x = '' ,  y = value1, geom='boxplot')
ggplot (data = my.df, aes( x= '', y=value1 )) + geom_boxplot()
```


### Two Dimension Data
- In **boxplot**, use ~ to specify **y~x**, where **x** is grouping variable

```{r}
boxplot(data = my.df, value1~grp)
qplot  (data = my.df, x = grp , y = value1, geom='boxplot')
ggplot (data = my.df, aes(x=grp, y=value1)) + geom_boxplot()
```

### Two + One Dimension Data
- Extra dimension can be included in for **x-axis**
- In **boxplot**, use **+** to specify extra dimension
- In **qplot/ggplot**, use **interaction** to specify extra dimension

```{r, fig.width=8.05}
boxplot(data = my.df, value1~grp+dept)
```

```{r, fig.width=8.05}
qplot  (data = my.df, x=interaction(grp,dept) , y=value1, geom='boxplot')
ggplot (data = my.df, aes(x=interaction(grp,dept) , y=value1)) + geom_boxplot()
```

## Pie Chart
