```{r 09-init, echo=FALSE}
## global chunk option ##
knitr::opts_chunk$set(echo=TRUE, message=FALSE, fig.width=9, fig.height=2.5, fig.show='hold') 
```

# Clustering Analysis

- Cluster analysis is a data-reduction technique designed to uncover subgroups of observations within a dataset  
- It reduce a large number of observations to a much smaller number of clusters or types  
- A cluster is defined as a group of observations that are more similar to each other than they are to the observations in other groups  

## Library

Below summarizes all packages, their functions and purposes used in this **Clustering Analysis** chapter.  

|   | Package           | Function      | Purpose                                                                                                 |
|---|-------------------|---------------|---------------------------------------------------------------------------------------------------------|
| 1 | Base - R          | dist          | Calculate distance between data points with methods: euclidean, maximum, cenberra, minkowski, manhattan |
|   |                   | scale         | Scale data (minus mean, div by SD)                                                                      |
|   |                   | hclust        | Build hirerchical cluster model (no cutting)                                                            |
|   |                   | kmeans        | Build k-means cluster model                                                                             |
| 2 | factoextra        | fviz_nbclust  | Optimum number of cluster (K) visual analysis, methods: wss, silhoutte                                  |
|   |                   | hcut          | Build hirerchical cluster model (with cutting)                                                          |
|   |                   | fviz_dend     | Visualize h-cluster model in dendrogram graph  as ggplot object                                         |
|   |                   | fviz_cluster  | Visualize data points with cluster grouping as ggplot object                                            |
| 3 | NbClust           | NbClust       | 30 indices to analyze optimal number of cluster, K                                                      |
| 4 | flexclust         | randIndex     | Agreement measure for two cluster results                                                               |

```{r, results='markdown', collapse=TRUE}
library(flexclust)
library(factoextra)
library(NbClust)
```

## Application


Finding groups within data and distinguishing observation in your data set that are similar from those that are different. The objective is to find the right balance between similarities and differences. On the one hand we want to treat similar cases in a similar way to benefit from economies of scale.  

Finding groups within data will allow you to allocate your effort more efficiently:  

- Leveraging synergies between cases that are similar  
- Allocating specific case houses to different cases if needed in order to maximize your effectiveness  


Example Usages are:  

- **Business** : researchers use cluster analysis for customer segmentation. Customers are arranged into clusters based on the similarity of their demographics and buying behaviours. Marketing campaings are then tailored to appeal to the groups  
  
  
- **Human Resource** : you're concerned by the number of employees leaving the firm. As an HR manager, you want to retain your best employees within the company but you cannot follow-up with each one of them too often as that would be very time consuming.  Instead, you may rely on an HR analytic solution to understand what are the most frequent situations explaining why an employee decides to leave, with variables like number of completed projects for the past 12 months/utilization, age, last project evaluation, time spent in company, position, number of newborn, etc.   

- **Psychological**: researchers cluster data on the symptoms and demographics of depressed patients, seeking to uncover subtypes of depression, with the hope of finding more effective targeted treatments and a better understanding of the disorder.
  
- **Medical**: researchers use cluster analysis to help catalog gene-expression patterns obtained from DNA microarray data. This can help them to understand normal growth and development and the underlying causes of many human diseases

- **Information Retrieval**: The world wide web consists of billions of Web pages, and the results of a query to a search engine can return thousands of pages. Clustering can be used to group these search results into a small number of clusters, each of which captures a particular aspect of the query. For example, a query of "movie" might return Web pages grouped into categories such as reviews, trailers, starts and theaters. Each category (Cluster) can be bnorken into subcategories (sub-clusters), producing a hierachical structure that further assists a user's exploration of the query results.

## Sample Data

Sample data used in this chapter emulate two dimensional data points with three groups with clear grouping when visualize. 

```{r fig.width=9, fig.height=4, fig.show='asis'}
set.seed(1234)
my.df = data.frame(
  id = paste('ID_', 1:18, sep = ''),
  grp = c(rep('G1', 6), rep('G2', 6), rep('G3', 6)),
  value1 = c( round(rnorm(6, mean = 10,  sd = 3)),
              round(rnorm(6, mean = 10, sd = 3)),
              round(rnorm(6, mean = 30, sd = 3))),
  value2 = c( round(rnorm(6, mean = 10, sd = 3)),
              round(rnorm(6, mean = 20, sd = 3)),
              round(rnorm(6, mean = 20, sd = 3))),
  stringsAsFactors = F
)
rownames(my.df) = my.df[,1]
my.df
plot(my.df$value1, my.df$value2)
```

```{r}
spl = sample(1:nrow(my.df), 0.7*nrow(my.df))
train = my.df[spl,  ]
test  = my.df[-spl, ]
```

## General Steps

1. **Choose appropriate attributes**  
    + This is the most important steps. 
    + Choose attributes that actions can be taken upon
    + A sohisticated cluster analysis can't compensate for a poor choice of variables  <br><br>

2. **Scale Data**  

    **When NOT to scale **  

    If you have attributes with a well-defined meaning. Say, latitude and longitude, then you should not scale your data, because this will cause distortion  

    **When To Scale **  

    + If you have mixed numerical data, where each attribute is something entirely different (say, shoe size and weight), has different units attached (lb, tons, m, kg ...) then these values aren't really comparable anyway; z-standardizing them is a best-practise to give equal weight to them  

    + If variables vary in range, then the variable with the largest value will have the greatest impact on result. This is undesirable  

    + Therefore data must be scaled so that they can be compared fairly  

    **Methods of Scaling**  

    Popular scaling methods are:  
      + Normalize to mean=0 and sd=1  
      + Divide by Max  
      + Minus min, divide by Min-Max range  <br><br>  

3. **Screen for Outliers**  
    + Outliers can distort results. Screen to remove them  <br><br>
    
4. **Calculate Data Point Distances**
    + Popular measure of distance between two data point is Euclidean distance
    + Others are Manhattan, Canberra, Asymmetric Binary, Maximum and Minkowski also available  <br><br>
    
5. **Chosoe a Clustering Alrorithm, and Inter-Custer Distance Method**
    

6. **Try few Clustering Solutions**

    + Decide the best clustering algorithm, cluster distance method and **number of cluster, K**  
    + Use `NbClus` as a tool to guide choosing K (number of cluster)  

7. **Visualize the result**
    + Visualization can help you determine the meaning and usefulness of the cluster solution  
    + **Hierarchical** clustering are usually presented as a dendrogram  
    + **Partitioning** results are typically visualized using a bivariate cluster plot  <br><br>

8. **Intepret the Cluster**
    + Once a cluster solution has been obtained, you must interpret (and possibly name) the clusters
    + What do the observations in a cluster have in common? 
    + How do they differ from the observations in other clusters? 
    + This step is typically accomplished by obtaining summary statistics for each variable by cluster
    + For continuous data, the mean or median for each variable within each cluster is calculated. 
    + For mixed data (data that contain categorical variables), the summary statistics will also include modes or category distributions  <br><br>
    
9. **Validate Result**
    + Validating the cluster solution involves asking the question:
        + “Are these groupings in some sense real, and not a manifestation of unique aspects of this dataset or statistical technique?”   
        + If a different cluster method or different sample is employed, would the same clusters be obtained?  
        + If actual grouping data is known, run randIndex to measure the degree of agreement  
        + The fpc, clv, and clValid packages each contain functions for evaluating the stability of a clustering solution (not discussed here)  

## Distance Algorithm

- The choice of an appropriate metric will influence the shape of the clusters  
- As some elements may be close to one another according to one distance and farther away according to another   
- For example, in a two dimensional data, the distance between the point (1,1) and the origin (0,0) can be 2 under **Manhattan distance**, $\sqrt{2}$ under **Euclidean distance**, or 1 under **Maximum distance**.

**`dist`** is used to measure distance for all **numeric**  elements in dataframe or matrix. Supplying non-numeric columns for `dist` will incur warning.

> **`dist`**`( x, method = )   default method = 'euclidean'`  
> $\quad$ `method = 'euclidean', "maximum", "manhattan", "canberra", "binary" or "minkowski"`  

### Computation Overhead

Computation power are heavy for distance calculation for below reason:  

- For N observations, **`dist`** will generate N(N-1)/2 distance matrices  
- The more variables in the data, the longer it takes to calculate  

For large datasets, it is likely feasible to calcualte distance as it is take too long and too much memory.  

### Euclidean Distance

$$Euclidean-d(p,q) = \sqrt{\sum_{i=1}^n (p_i-q_i)^2} \quad,n = dimension$$  

- The Euclidean distance is a distance measure between two points or or vectors in a two- or multidimensional (Euclidean) space **based on Pythagoras' theorem**   
- The distance is calculated by taking the square root of the sum of the squared pair-wise distances of every dimension 

Below command measures distance for  numeric columns of **all data points** in my.df, using **euclidean** algorithmn.  

```{r}
train.scaled = scale(train[,3:4])    # Z-Score scaling for Train Data
test.scaled = scale(test[,3:4])      # Z-Score scaling for Test Data
d.euclidean = dist( train.scaled )   # Euclidean distance
round (d.euclidean,1)
```

### Manhattan Distance

$$Manhattan - d(p,q) = \sum_{i=1}^n |p_i-q_i| \quad,n = dimension$$  

+ The Manhattan distance (sometimes also called **Taxicab** distance) metric is related to the Euclidean distance
+ But instead of calculating the shortest diagonal path ("beeline") between two points, it calculates the distance based on gridlines  

Below command measures distance for  numeric columns of **all data points** in my.df, using **manhattan** algorithm.  

```{r}
train.scaled = scale(train[,3:4])    # Z-Score scaling for Train Data
test.scaled = scale(test[,3:4])      # Z-Score scaling for Test Data
d.manhattan = dist( train.scaled, method='manhattan')
round (d.manhattan, 1)
```

### Maximum Distance

$$d(x,y)= sup|x_j - y_j|, 1≤ j ≤ d$$

### Canberra Distance

$$\sum_{j=1}^{d}|x_j - y_j|) / (|x_j|+|y_j|)$$



### Minkowski Distance

The Minkowski distance is a generalized form of the **Euclidean distance** (if m=2) and the **Manhattan distance** (if m=1).  

$$\left(\sum_{i=1}^n |p_i-q_i|^p\right)^{1/m}$$


## Optimum Number of Clusters (K)  

There are three (3) popular methods for determining the optimal number of clusters.  

1. Elbow Method
    Applicable for partioning clustering, such as k-means
    
2. Average Silhoutte Method  


3. Gap Statistics  (not discussed here)


There is no guarantee that they will agree with each other. In fact, they probably won’t. However, use this as a guidine and test few highest criteria score to determinee final number of cluster.  

### Elbow Method 

#### Elbow Concept

The objective of partitioning clustering (such as K-Mean) is to define clusters such that the total intra-cluster variation (known as total within-cluster variation or total within-cluster sum of square, wss) is minimized.

#### Elbow Algorithm

1. Run K-mean clustering algorithm for K=1 to n  
2. For each K, calculate the within-cluster-sum-of-square (wss)  
3. Plot the curve of wss against the number of clusters K  
4. The location of bend (knee) in the plot is generally considered as the indicator of the appropriate number of clusters  

When the WSS value stop decreasing significantly (at the knee), then the number of clusters probably had reached its optimum. Although this approach is heuristic, it still provide a good guideline for K selection.  

#### Elbow Codes (for K-mean) - Do It Yourself!

The method presented here does not require any external library !  However, it requires writing a funciton to calculate WSS and plot the results.  

**Define the The Algorithmn**  

```{r}
# Algorithmn: Compute k-means and plot wss for k=2 to k=15
wssplot = function(data, nc=15, seed=1234){
            wss <- (nrow(data)-1)*sum(apply(data,2,var))
            for (i in 2:nc) {
              set.seed(seed)
              wss[i] <- sum(kmeans(data, centers=i)$withinss)
            }
            plot(1:nc, wss, type="b", 
                xlab="Number of Clusters (K)",
                ylab="Total Within Groups Sum of Squares")
            wss
}
```

**Run The Code**  

If number of observations <=nc(default 15), specify smaller nc.  

```{r, collapse=TRUE, fig.height=4, fig.show='hold'}
wssplot(train.scaled, nc=6)
abline(v=3, lty=2)     # mark the optimum K after facts 
```

The wssplot above indicates that there is a **distinct drop** in the within-groups sum of squares when moving **from 1 to 3 clusters**. After three clusters, this **decrease drops off**, suggestign that a three-cluster solution may be a good fit to the data.  

#### Elbow Codes - using `factoextra::fviz_nbclust,hcut`

+ `factoextra` combined functions to calculate 'silhoutte' and output `ggplot` object  
+ For k-mean wss analysis, `kmeans` helper function from base-R is required  
+ For pam wss analysis, `cluster:pam` helper function is required  
+ For h-cluster wss analysis, `hcut` helper function by its own library is used. Somehow base-R `hclust` is not supproted  

```{r, echo=FALSE}
par(mfrow=c(1,3))
```

```{r, fig.height=4, fig.width=2.65, fig.show='hold'}
fviz_nbclust(train.scaled, kmeans, method = "wss") + labs(subtitle='kmeans')
fviz_nbclust(train.scaled, pam,    method = "wss") + labs(subtitle='pam')
fviz_nbclust(train.scaled, hcut,   method = "wss") + labs(subtitle='hcut') +
  geom_vline(xintercept = 3, linetype = 2)
```

### Average Silhoutte Method

#### Average Silhoutte Concept

Average silhouette method computes the average silhouette of observations for different values of k. The **optimal number** of clusters k is the one that **maximize the average silhouette** over a range of possible values for k (Kaufman and Rousseeuw [1990]).

Silhouette analysis can be used to study the **separation distance between the resulting clusters**. The silhouette plot displays a measure of how close each point in one cluster is to points in the neighboring clusters and thus provides a way to assess parameters like number of clusters visually. This measure has a range of [-1, 1].

Silhouette coefficients (as these values are referred to as) **near +1 indicate that the sample is far away from the neighboring clusters**. A value of 0 indicates that the sample is on or very close to the decision boundary between two neighboring clusters and negative values indicate that those samples might have been assigned to the wrong cluster.

#### Average Silhoutte Algorithm

1. Compute clustering algorithm (e.g., k-means clustering) for different values of k  
2. For each k, calculate the average silhouette of observations (avg.sil)  
3. Plot the curve of avg.sil according to the number of clusters k  
4. The location of the maximum is considered as the appropriate number of clusters  

#### Average Silhoutte Code - `factoextra:fviz_nbclust`

Example code below shows silhoute analysis for kmeans, pam and h-cluster:  

+ `factoextra` combined functions to calculate 'silhoutte' and output ggplot object   
+ For k-mean silhoutte analysis, `kmeans` helper function from base-R is required  
+ For pam silhoutte analysis, `cluster:pam` helper function is required  
+ For h-cluster silhoutte analysis, `hcut` helper function by its own library is used. Somehow base-R `hclust` is not supproted  

```{r, echo=FALSE}
par(mfrow=c(1,3))
```

```{r, fig.height=4, fig.width=2.65, fig.show='hold'}
fviz_nbclust(train.scaled, kmeans, method = "silhouette") + labs(subtitle='kmeans')
fviz_nbclust(train.scaled, pam,    method = "silhouette") + labs(subtitle='pam')
fviz_nbclust(train.scaled, hcut,   method = "silhouette") + labs(subtitle='hcut')
```

### NbClust Package (with 30 Indices)

`NbClust` package offers numerous 26 indices for determining the best number of clusters in a cluster analysis.  

+ Multiple indices are computed  **simultaneously** - a clear advantage  
+ Paramter `index='all'` will utilize all indices to evaluate the optimum number of clusters   
+ `Nbclust` returns a list that contains all evaluation statistic based on the indices used  
+ Results of the evaluation is stored in `Best.nc` vector  
+ Using `table` and `barplot` is best way to visualize the result of best K  

**Supported Indices are**  

+ kl, ch, hartigan, ccc, scott, marriot, trcovw, tracew, friedman, rubin, cindex, db, silhouette, duda, pseudot2, beale, ratkowsky, ball, ptbiserial, gap, frey, mcclain, gamma, gplus, tau, dunn, hubert, sdindex, dindex, sdbw  
+ **'all'** (all indices except GAP, Gamma, Gplus and Tau)  
+ 'alllong' (all indices with Gap, Gamma, Gplus and Tau included)  
 
> **`NbClust( data=, diss=NULL, distance='euclidean', min.nc=2, max.nc=15,`**  
> **`method=NULL, index='all', alphaBeale=0.1)`**  
> $\quad$ `data = matrix or dataframe`  
> $\quad$ `diss = dissimilarity matrix, if not NULL, then distance should be NULL`  
> $\quad$ `distance = "euclidean", "maximum", "manhattan", "canberra", "binary", "minkowski" or "NULL"`  
> $\quad$ `min.nc = minimum number of clusters`  
> $\quad$ `max.nc = maximum number of clusters`  
> $\quad$ `method = "ward.D", "ward.D2", "single", "complete", "average", "mcquitty", "median", "centroid", "kmeans"`  
> $\quad$ `index = 'all' to use all indices for evaluation`  

`NbClust` **output** an object with below values:  

> **`Best.nc`** `: Best number of clusters proposed by each index and the corresponding index value`  
> **`Best.partition`**` : vector of cluster group for every observation`   

#### Run The Code

Run `NbClust` for `average (h-clustering)` and `kmeans` method.  

```{r, warning=FALSE, results='hide', fig.show='hide', collapse=TRUE}
nbc.hclust = NbClust(train.scaled, distance="euclidean", min.nc=2, max.nc=8, method="average")
nbc.kmeans = NbClust(train.scaled,                       min.nc=2, max.nc=8, method="kmeans")
```

#### Visualize The Result

**Visualize using  Base-R**  

As output `Best.nc[1,]` shows, majority indices favor three (3) clusters.  

```{r, echo=FALSE}
par(mfrow=c(1,2))
```

```{r, warning=FALSE, fig.height=3.5, fig.width=4, collapse=TRUE}
table( nbc.hclust$Best.n[1,] )
barplot( table(nbc.hclust$Best.n[1,] ),
  xlab="Numer of Clusters", ylab="Number of Criteria",
  main="Number of Clusters Chosen by 26 Criteria\nh-cluster")

table( nbc.kmeans$Best.nc[1,] )
barplot( table(nbc.kmeans$Best.nc[1,] ),
  xlab="Numer of Clusters", ylab="Number of Criteria",
  main="Number of Clusters Chosen by 26 Criteria\nkmeans")
```

**Visualize using  `factoextra::fviz_nbclust()`**  

Single function `fviz_nbclust()` from `factoextra` library will use value in **NbClust** object to visualize the optimal cluster number.  `fviz_nbclus` output `ggplot` object, hence can be easily customized.  

```{r, warning=FALSE, fig.height=3.5, fig.width=4, collapse=TRUE, results='hide'}
fviz_nbclust(nbc.hclust) + labs(subtitle='H-Cluster')
fviz_nbclust(nbc.kmeans) + labs(subtitle='K-Means')
```


## Clustering Algorithm Compared

|   | Description           | h-cluster                                     | k-means                                   |
|---|:----------------------|:----------------------------------------------|:------------------------------------------|
| 1 | Computation Time      | Fast.  Linear to number of observation        | Slow: Quadradric to number of observation |
| 2 | Initial K needed      | No                                            | Yes                                       |
| 3 | Fine Tuning           | Experiment with different method of Linkage   | Experiment with different K centroids     |
| 4 | Perform Well in       | Hierachical Nature Data Set                   | Spherical Data Points                     |
| 5 | Perform Bad  in       | Large data sets                               | U-Shape, Outliers                         |
| 6 | Unique Advantages     | Good for hirechical discovery                 |                                           |
| 7 | R Library             | Base R,  factoextra                           | Base R                                    |

## Hierarchical Clustering 

+ Hierarchical clustering is a widely used data analysis tool  
+ The idea is to build a binary tree of the data that successively merges similar groups of points  
+ Number of clusters (K) is required as import  
+ It is an unsupervised learning  

### Clustering Algorithm

This is how Hierarchical Clustering works:  

    1. Initially, put each data point in **its own cluster**  
    2. **Calucate the distances** between each cluster and all other clusters (inter-cluster distance)  
    3. **Combine the two clusters** with the **smallest distance** - This reduce cluster number by one  
    4. Repeat step (2) and (3) until all clusters have been **merged into single cluster**  <br>

### Inter Cluster Distance Method

Once distance for all data points has been measured, decide which of the five (5) methods below to measure distance between clusters:  

1. **Single Linkage**:  
   Shortest distance among all data points betweentwo clusters    
2. **Complete Linkage (common)**:  
   Longest distance among all data points between two clusters   
3. **Average Linkage (common)**:  
   Average distance of all points between two clusters
4. **Centroid**:  
   Find the centroid of each cluster and calculate the distance between centroids between two clusters  

Please note that the **Inter Cluster Distance Method** above uses **Distance Algorithmn** such as 'euclidean', "maximum", "manhattan", "canberra", "binary" or "minkowski" to calculate actual distance value.  

### Run The Code

Generally, step (A), (B) and (C) are followed for H-clustering analysis.  

**A. Before running H-Clustering**  

1. Filter Data (choose only numeric columns)  
2. Scale Data (if required)  
3. Calculate distance, choose a distance algorithm

**B. Performing H-Clustering**  

1. Build H-Cluster model,  require input of **inter-cluster distance method**  
2. Derive cluster by **cutting** into K number of clusters  

**C. Visualize**

1. Display frequency, number of observations for each cluster group  
2. Plot dendrogram, superimpose cluster group into the plot  

#### Using Base-R Utilities

**Filter, Scale, Calculate Distance**  

> **`dist (d, method ='euclidean')`**  
> $\quad$ `x      =  matrix or data.frame`  
> $\quad$ `method = 'euclidean', "maximum", "manhattan", "canberra", "binary"`  

```{r}
train.scaled = scale(train[,3:4])   # scale data
d.euclidean = dist( train.scaled )  # calculate distance
```

**Build H-Cluster Model, Cutting into clusters**  

> **`hclust (d, method ='complete')`**  
> $\quad$ `d      = distance calculated using dist()`  
> $\quad$ `method = 'single', 'complete', 'average', 'centroid' as cluster distance method`  

> **`cutree (obj, k=NULL, h=NULL)`**  
> $\quad$ `obj = hclust object`  
> $\quad$ `k   = integer, total number of expected groups`    

```{r, collapse=TRUE}
fit.average  = hclust (d.euclidean, method='average')  # build the model
clusters = cutree (fit.average, k=3)                   # derive the clusters
clusters
```

**Display frequency table, Visualize with dendogram, superimpose cluster group**  

```{r echo=FALSE}
par(mfrow=c(1,2))
```

```{r fig.height=4, fig.width=4, fig.show='hold'}
barplot( table(clusters), xlab='Cluster Group Number', ylab='No. of Observations' ) 
plot( fit.average, main='HCluster, Average-Linkage\n3 Clusters Solution' )
rect.hclust( fit.average, k=3, border = 2:5 )
```


#### Using `factoextra` Package

`factoextra` provides single function `hcut` to **scale**, **calcuate distance** and **cutting** into cluster groups. Which is **handy**.  

> `hcut(x, k = 2, hc_method = "ward.D2", hc_metric = "euclidean", stand = FALSE, graph = FALSE)`  
> $\quad$ `x         = matrix or dataframe`  
> $\quad$ `k         = number of clusters to cut`  
> $\quad$ `hc_method = inter-cluster distance method: ward.D, ward.D2, single, complete, average`  
> $\quad$ `hc_metric = distance calc method: euclidean, manhattan, maximum, canberra, binary, "minkowski`  
> $\quad$ `stand     = TRUE: scale x with z-score, FALSE: not to scale x`  

`hcut` output below useful values (not all displayed):  

> $\quad$ `data    = original data (if stand=FALSE), scaled data (if stand=TRUE)`  
> $\quad$ `nbclust = number of clusters`  
> $\quad$ `cluster = cluster group number assigned for each observation`  
> $\quad$ `size    = frequency vector, number of observations for each cluster`  
> $\quad$ `method  = inter-cluster distance method applied`  
> $\quad$ `dist.method  = distance method applied`  

**Filter, Scale, Calculate Distance, Build H-Cluster Model, Cutting into Clusters**  

```{r}
fit.hc = hcut(train[,3:4], k=3, hc_method='average', hc_metric='euclidean', stand = TRUE)
```

**Display frequency table, Visualize with dendogram, superimpose cluster group**  

```{r echo=FALSE}
par(mfrow=c(1,2))
```

```{r fig.height=4, fig.width=4, fig.show='hold'}
barplot( table(fit.hc$cluster), xlab='Cluster Group Number', ylab='No. of Observations' ) 
fviz_dend(fit.hc, rect = TRUE, rect_border = 'red', cex = 0.5, lower_rect = -0.5, horiz = T)
```


## K-Mean Clustering

+ **K-mean** is the **most common partitioning** clustering algorithm  
+ Partitioning means data points need to be initally partioned into few clusters to start the process with  
+ The other partitioning clustering method is **Medoids**  

### Clustering Algorithm

1. **Define K number of centroids (data points)**  

2. **Cluster Assignment**  
   Each observation is assigned to the nearest centroid, using **euclidean** distance

3. **Update Centroids**  
   After all observations had been assigned to the centroids, a new centroids is calculated  

4. **Repeat step (2) - (3) until convergence**  
   Convergence means none of the observations changed  cluster membership  


### Run Code: Calc KMeans

Generally, step (A), (B) and (C) are followed for K-Mean clustering.  

#### General Steps

**A. Before running K-Mean**  

1. Filter Data (choose only numeric columns)  
2. Scale Data (if required)  

**B. Performing K-Mean**  

1. Build K-Means Cluster model,  require input  
    + Number of initial centers (clusters K)
    + K-Means Algorithmn 
    + Number of tries to seed random centers, before choosing the best model

**C. Visualize**  

1. Display frequency, number of observations for each cluster group  
2. Plot graph, superimpose cluster group into the plot  

#### Run K-Means

> **`kmeans( x, centers, iter.max=10 nstart=1, algorithmn='Hartiga-Wong' )`**  
> $\quad$ `x = matrix or dataframe`  
> $\quad$ `centers   = number of centroids`  
> $\quad$ `nstart    = how many times to randomly try seeding centroids`  
> $\quad$ `algorithm = "Hartigan-Wong", "Lloyd", "Forgy", "MacQueen"`  
> $\quad$ `iter.max  = maximum number of iterations allowed`  

**Choosing `nstart`**

- **nstart=25** is a good number to start with generally  
- It will run kmeans 25 times with different random centers, and choose the one with **lowest within cluster sum of square**  
- nstart is a **hard stop**  

**Choosing `iter.max`**  

- kmeans will **iterate** until convergence  
- Convergence is reach when **errors (measured by sum of square) has been minimized**, in other words, distance between clusters stablized  
- If **iter.max is too low**, kmeans process will stop before reaching minimum error; hence not optimized  

**Calculate K-Means**  
```{r, collapse=TRUE}
kmc = kmeans(train.scaled, 3, nstart = 25)
```

#### K-Means Output

`kmeans` output below useful values:  

> **`cluster`** ` : cluster assignment for all observations`  
> **`centers`** `: Centre values (of each dimensions) for each cluster`  
> **`withinss`** `: Total Sum of Squares Within For All Cluster`  
> **`betweennss`** `: Total Sum of Squares Between clusters`  
> **`size`** `: Number of observations for each cluster`  

```{r, collapse=TRUE}
str(kmc)
```

```{r, collapse=TRUE}
kmc$cluster
```

```{r, collapse=TRUE}
kmc$withinss
```

```{r, collapse=TRUE}
kmc$size
```

#### Visualizing K-Mean Cluster

```{r, fig.height=5}
fviz_cluster( kmc, data = train.scaled, 
  geom = "point", stand = FALSE, ellipse.type = "norm")
```

### Run Code: Build Model

####Build The Model
- Kmeans object produced above contain information sufficient to build a model  
- We build the model with training the kmean object with training data  
- Unfortunately, there is no single step to calculate kmeans and train the model  
```{r}
fit.kcca = as.kcca(kmc, train.scaled)
```

####Prediction
As in any other modeling, we can use `predict` to derive clustering groups on test data  
```{r}
pred.kcca = predict(fit.kcca, newdata=test.scaled)
pred.kcca
```

## Cluster Agreement

### Compare Clustering Results

We want to compare clusters in two scenarios:  

- Compare result from cluster algorithm with actual cluster group; measure how close is our model to actual  
- Comapre result from two cluster algorithms; analyze corresponding cluster number between two algorithmn (if any)  

Comparing allow us know which cluster from one algorithm correspond to which cluster in another algorithm; because the cluster number derived from the alrogithmn does not represent any sequence or meaning.  

### Visualize Cluster Differences
- Construct a table of cluster groups to be compared  
- If both agree to each other, the observations matched will be high  

**Compare result with actual grouping**
```{r collapse=TRUE}
table(train$grp, clusters)
```
**Compare results of two algorithms**
```{r collapse=TRUE}
table(kmc$cluster, fit.hc$cluster, dnn=c('kmeans','hcluster'))
```

### Agreement Measurement
- The **adjusted Rand index** provides a measure of the agreement between two sets of grouping data, adjusted for chance  
- It ranges from -1 (no agreement) to 1 (perfect agreement)  
- **`flexclust::randIndex`** is the function to calculate this index, it takes **2-dimensional cross-tabulation** of cluster assignments as input  
**Compare agreement with actual grouping**
```{r collapse=TRUE}
randIndex( table(train$grp, clusters) )
```
 **Compare agreement with two algorithms**
```{r collapse=TRUE}
randIndex( table(kmc$cluster, fit.hc$cluster) )
```
