---
title: "Survival Analysis"
output: html_document
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=9)
```



# Survival Analysis

## Library Used

```{r}
library(survival)
```

## Sample Data

```{r results='markdown'}
main.data = read.csv('./datasets/surv_maintenance.csv')
str(main.data)
head(main.data)
summary(main.data)
```

## The Code

**Create the Model**  

Step 1 - Use survival::Surv to create an object that ties time (lifetime) to event (broken). This object simple simply combines both time and event variables together in a matrix.  

Step 2 - run the modelling with the time-event object create above, specify the parameters and distribution (example below use normal distribution).  

```{r}
dependantVar = Surv(main.data$lifetime, main.data$broken)
main.fit = survreg(dependantVar~pressureInd+moistureInd+temperatureInd+team+provider, 
                   dist="gaussian", 
                   data=main.data)
summary(main.fit)
```

Use the model to predict lifetime before break (labeled as Ebreak below).  Then calculate the remaing lifetime (remaingLT) as predicted lifetime before break - current lifetime.

```{r}
Ebreak = predict(main.fit, data=main.data,type='quantile',p=0.5)
remainingLT = Ebreak - main.data$lifetime
```

```{r}
main.data.forecast = data.frame (
  lifetime = main.data$lifetime,
  broken = main.data$broken,
  Ebreak,
  remainingLT
)
```

Put all into one data.frame.  Sort the remainingLT with lowest first, so that the priority to pay attention to are top to down.  

```{r}
head(main.data.forecast[order(main.data.forecast$remainingLT, decreasing = FALSE),])
```


## Performance Evaluation

Performance can be measured by analysis **how well the model predict the lifetime of 'already happened event'**, which is the training data.  Subset the training data.  

```{r}
eval.data = subset ( main.data.forecast, broken==1)
```

Measure **correlation** of the predicted and actual.  

```{r, collapse=TRUE}
cor(eval.data$Ebreak,eval.data$lifetime)
```

**Visualize** the fitness of the model.  

```{r}
plot(eval.data$Ebreak, eval.data$lifetime)
```
